---
title: "R Notebook"
output: html_notebook
---

```{r}

library(MASS)
library(tidyverse)
library(here)
library(ggplot2)
library(lubridate)
library(GGally)
library(caret)
library(Amelia)

```

create clean data
```{r}
source(here("scripts/primary_data_cleaning.R"))
```

load clean data
```{r}
clean_data <- read_csv(here("clean_data/primary_clean_data.csv"), show_col_types = FALSE)
```

basic delay data
```{r}
clean_data %>%
group_by(origin_airport_name) %>% 
  ggplot() +
  aes(x = origin_airport_name, fill = dep_delay_bool) +
  geom_bar() +
  labs(x = "NYC Airport",
y = "Flights",
title = "Flights Per NYC Airport",
fill = "Delay") +
  scale_fill_manual(labels = c("on_time", "delayed"), values = c("steelblue", "maroon")) +
  theme_bw()


```
significant delays all round, more delays in newark but more flights total.
there seems to be no imbalance

```{r}
clean_data %>%
  select(origin_airport_name, dep_delay_bool) %>%
  na.omit() %>%
group_by(origin_airport_name) %>%
  summarise(percentage_delay = sum(dep_delay_bool)/n()*100)
```



```{r}
clean_data %>% 
  filter(origin_airport_name == "Newark Liberty International Airport") %>%
  ggplot() +
  aes(x = POSIXct_local_time, y = dep_delay) +
  geom_bin2d() + 
  coord_cartesian(ylim = c(-100, 1600)) +
  labs(title = "NEWARK AIRPORT", x = "Date", y = "Total Delay")

clean_data %>% 
  filter(origin_airport_name == "John F Kennedy International Airport") %>%
  ggplot() +
  aes(x = POSIXct_local_time, y = dep_delay) +
  geom_bin2d()+ 
  coord_cartesian(ylim = c(-100, 1600)) +
  labs(title = "JFK AIRPORT", x = "Date", y = "Total Delay")

clean_data %>% 
  filter(origin_airport_name == "La Guardia Airport") %>%
  ggplot() +
  aes(x = POSIXct_local_time, y = dep_delay) +
  geom_bin2d()+ 
  coord_cartesian(ylim = c(-100, 1600)) +
  labs(title = "GUARDIA AIRPORT", x = "Date", y = "Total Delay")
```


```{r}
clean_data %>% 
  filter(origin_airport_name == "Newark Liberty International Airport") %>%
  ggplot() +
  aes(x = POSIXct_local_time, y = dep_delay) +
  geom_point() + 
  coord_cartesian(ylim = c(-100, 1600)) +
  labs(title = "NEWARK AIRPORT", x = "Date", y = "Total Delay")

clean_data %>% 
  filter(origin_airport_name == "John F Kennedy International Airport") %>%
  ggplot() +
  aes(x = POSIXct_local_time, y = dep_delay) +
  geom_point()+ 
  coord_cartesian(ylim = c(-100, 1600)) +
  labs(title = "JFK AIRPORT", x = "Date", y = "Total Delay")

clean_data %>% 
  filter(origin_airport_name == "La Guardia Airport") %>%
  ggplot() +
  aes(x = POSIXct_local_time, y = dep_delay) +
  geom_point()+ 
  coord_cartesian(ylim = c(-100, 1600)) +
  labs(title = "GUARDIA AIRPORT", x = "Date", y = "Total Delay")
```


```{r}
clean_data %>%
group_by(airline_name) %>% 
  ggplot() +
  aes(y = airline_name, fill = dep_delay_bool) +
  geom_bar()

```


delay analysis over time
```{r}
clean_data %>%
  ggplot() +
  geom_histogram(aes(x = POSIXct_local_time, fill = dep_delay_bool), bins = 24) +
  labs(x = "DATE",
y = "Flights",
title = "2017_NYC_Flights",
fill = "Delay") +
  scale_fill_manual(labels = c("on_time", "delayed"), values = c("steelblue", "maroon")) +
  theme_bw()
```
delays seem more frequent from march to august

by day analysis
```{r}
clean_data %>%
    group_by(day = weekdays(POSIXct_local_time), delays =dep_delay_bool) %>%
    summarize(count = n()) %>%
    mutate(days = factor(day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  
ggplot(aes(x = days, y = count, fill = delays)) + geom_bar(stat='identity', position = "dodge")

```
no obvious increase in delays by day but a reduced overall volume on saterday.



```{r}
clean_data %>%
    group_by(hour = hour(POSIXct_local_time), delays =dep_delay_bool) %>%
    summarize(count = n()) -> by_day

by_day%>%
ggplot(aes(x = hour, y = count, fill = delays)) + geom_bar(stat='identity')
```

flights have a higher cancellation rate after midday and delays became more frequent after 3pm.



```{r}
clean_data %>%
    select(dep_delay_bool, POSIXct_local_time) %>%
    na.omit() %>%
    group_by(hour = hour(POSIXct_local_time)) %>%
    summarize(delays = sum(dep_delay_bool)/n()*100) -> by_day_line

by_day_line%>%
ggplot(aes(x = hour, y = delays)) + geom_bar(stat='identity')
```


```{r}
sapply(names(clean_data), function(x){mean(is.na(clean_data[[x]]))})
```

```{r}
missmap(clean_data,  rank.order = TRUE)
```



```{r}
clean_data_f = clean_data %>%
  select("flight",
       "tailnum",
       "airline_name",
       "POSIXct_local_time",
       
       "manufacturer",
       "manufacture_year",
       "model", 
       "type", 
       "seats",
       "engine",
       "engines",
       "origin_airport_name",
       "origin_airport_alt",
       "origin_airport_tzone",
       "wind_dir",
       "wind_speed",
       "wind_gust",
       "visib",
       
       "sched_dep_time",
       "dep_time",
       "dep_delay",
       "dep_delay_binary",
       "dep_delay_bool",
       "cancelled",
       
       "air_time",
       "distance"
       )
```



```{r}
lm1 <- lm(dep_delay~wind_speed, clean_data_f)

plot(lm1)

summary(lm1)
```


```{r}
wind_clean_data <- clean_data_f%>%
  select(dep_delay_binary, wind_speed, visib, seats, engines, POSIXct_local_time) %>% na.omit() %>%
  mutate(month = month(POSIXct_local_time))

wind_clean_data %>%

  group_by(month) %>%
  summarise(wind_speed = mean(wind_speed)/max(wind_clean_data$wind_speed)*100, prcnt_flights_delayed = sum(dep_delay_binary)/length(dep_delay_binary)*100) %>%

  
  ggplot(aes( x = month)) +
  geom_bar(aes(x = month, y = prcnt_flights_delayed), stat = "identity") +
  geom_line(aes(x = month, y = wind_speed), stat = "identity") +
  scale_x_discrete(limits = 1:12) +
  labs(x = "DATE",
y = "% flights delayed & % windspeed",
title = "monthly delays & windspeed",
fill = "Delay") +
  scale_fill_manual(labels = c("on_time", "delayed"), values = c("steelblue", "maroon")) +
  theme_bw()

```






```{r}

wind_clean_data <- clean_data_f%>%
  select(dep_delay_binary, wind_speed, visib, seats, engines, POSIXct_local_time) %>% na.omit() %>%
  mutate(month = month(POSIXct_local_time))

index = sort(sample(nrow(wind_clean_data), nrow(wind_clean_data)*0.8))

train<-wind_clean_data[index,]
nrow(train)
test<-wind_clean_data[-index,]
nrow(test)

liniar_model <- glm(dep_delay_binary ~ wind_speed + visib + seats + engines, family = binomial(link = "logit"), data = train)
summary(liniar_model)

prediction <- predict(liniar_model, newdata = test, type = "response")
head(prediction)
prediction_binary <- factor(ifelse(prediction > 0.5, 1, 0))
plot(liniar_model)
confusionMatrix(prediction_binary, factor(test$dep_delay_binary))

with(summary(liniar_model), 1 - deviance/null.deviance)
```
Accuracy : 0.6181   due to skewing of data

0.005410407 r^2 variance explained by the model, very small.

Confusion Matrix and Statistics

          Reference
Prediction     0     1
         0 28317 17223
         1   984  1155


```{r}

index = sort(sample(nrow(wind_clean_data), nrow(wind_clean_data)*0.8))

train<-wind_clean_data[index,]
nrow(train)
test<-wind_clean_data[-index,]
nrow(test)

fit <- glm(dep_delay_binary ~ wind_speed, family = binomial(link = "logit"), data = train)

result <- predict(fit, newdata=test, type="response")

plot(fit)

summary(fit)

```


```{r}

clean_data_f %>%
  filter(dep_delay_binary = TRUE) %>%
  select(dep_delay, visib) %>%
  na.omit() %>%
  group_by(visib) %>%
  summarise(avg_delay_count = mean(dep_delay)) -> vis_data

vis_data

vis_data %>%

ggplot(aes(x = visib, y = avg_delay_count)) +
geom_point() + 
geom_smooth(method = "lm") +
labs(x = "visibility (miles)",
y = "avg delays (minutes)",
title = "visibility delay relationship")

```


```{r}
clean_data_f %>%
  filter(dep_delay_bool = TRUE) %>%
  select(dep_delay, seats) %>%
  mutate(seats = factor(seats)) %>%
  na.omit() %>%
  group_by(seats) %>%
  summarise(avg_delay_count = mean(dep_delay)) %>%
  ggplot() +
    geom_bar(aes(x = seats, y = avg_delay_count), stat = "identity")


clean_data_f %>%
    filter(dep_delay_bool = TRUE) %>%
  select(dep_delay, seats) %>%
  mutate(seats = factor(seats)) %>%
  na.omit() %>%
  group_by(seats) %>%
  summarise(count = n()) %>%
  ggplot() +
    geom_bar(aes(x = seats, y = count), stat = "identity")

```


